---
title: "Project Markdown"
author: "Nojoud Almohanna"
date: "10/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Developing a model to describe bacterial growth

The Bacterial growth data below obtained from a V. natriegens (a marine bacterium) experiment which aimed to examine the growth rate of V. natriegens in a laboratory.
The goal is to build discrete dynamical system models describing its population growth.
For more details see <https://mathinsight.org/bacteria_growth_initial_model>.

The bacteria population size was recorded every 16 minutes.

#### Read the data

```{r}
bacden <- read.csv("~/Desktop/CS 510/MidtermProject/ Bacterial density data.csv")
head(bacden)
```
#### Visualizing population density versus time

```{r}
plot(x = bacden$Time, y = bacden$Population.Density,col = 'black', bg = 'red', pch = 21)
```




It's clear that the density is increasing faster as the time progresses.

#### Calculating population changes

```{r}
counts = bacden$Population.Density
l = counts[-1]-counts[-length(counts)]
c <- round(l, 3)
c
```

#### Visualizing population change versus time

```{r}
plot(x=bacden$Time, y=c[1:length(bacden$Time)], col = 'black', bg = 'red', pch = 21)
```



The population change increases as a function of time.

## Fitting a linear model to population change versus density

```{r}
library(growthrates)
fit <- fit_easylinear(bacden$Time, l[1:length(bacden$Time)])
summary(fit)
```

#### Plotting population change versus population size

```{r}
plot(fit, log = "y") #log-scale
```
```{r}
coef(fit) # exponential growth parameters
```

```{r}
deviance(fit)  # residual sum of squares
```

```{r}
rsquared(fit)  # coefficient of determination
```

## Predicting the growth under dynamic conditions

In predictive microbiology mathematical models are usually divided in primary and secondary models.

- **Primary models** describe the variation of the population size against time. 

- **Secondary models** describe the relationship between the parameters of the primary model and the environmental conditions.


The **biogrowth** package can be used for simulating growth under dynamic environmental conditions using **predict_dynamic_growth()** function. This function combines primary and secondary growth models. It has 7 arguments:

- times: Numeric vector of time points for the calculations.
- env_conditions: A tibble describing the variation of the environmental conditions.
- primary_pars: A named list describing the model parameters of the primary model.
- secondary_models: A nested list defining the secondary model(s).
- ...: Additional arguments passed to the numeric solver of the differential equation.
- check: Whether to do some validity checks of model parameters (TRUE by default).
- formula: A one-sided formula describing the x variable in env_conditions.


I will consider two environmental factors: *temperature* and *pH*, defined using a tibble.

```{r}
library(ggplot2)
library(tibble)
my_conditions <- tibble(Time=c(0, 5, 80),
                         temperature = c(20, 30, 35),
                        pH = c(7, 6.5, 5))
```

#### Temperature plot

```{r}
ggplot(my_conditions) + geom_line(aes(x = Time, y = temperature))
```

#### PH plot

```{r}
ggplot(my_conditions) + geom_line(aes(x = Time, y = pH))
```



#### Define model parameters


```{r}
# For dynamic conditions,biogrowth uses the Baranyi growth model as primary model.

my_primary <- list(mu_opt = .9, # the maximum specific growth rate
             Nmax = 1e8, # the maximum growth rate
             N0 = 0.022, # the initial population size
             Q0 = 1e-3)
```

```{r}
sec_temperature <- list(model = "Zwietering",
                        xmin = 25,
                        xopt = 35,
                        n = 1)
```

```{r}
sec_pH <- list(model = "CPM",
               xmin = 5.5,
               xopt = 6.5,
               xmax = 7.5,
               n = 2)
```

```{r}
my_secondary <- list(
    temperature = sec_temperature,
    pH = sec_pH
    )
```

```{r}
my_times <- seq(0, 50, length = 1000)
```

```{r}
library(biogrowth)
dynamic_prediction <- predict_dynamic_growth(my_times,
                       my_conditions, my_primary,
                       my_secondary,
                       formula = . ~ Time)

```

#### Simulation

```{r}
dynamic_prediction$simulation
```
#### Plot the results
```{r}
plot(dynamic_prediction)
```


```{r}
# plot some environmental factor with add_factor
plot(dynamic_prediction, add_factor = "temperature")
```


```{r}
# Time to reach a given population size
time_to_logcount(dynamic_prediction, 6)
```

